---
title: "p8105_hw3_cj2793"
author: "Chenyu Jin"
date: "2024-10-08"
output: github_document
---
# Problem 1

## This problem uses the `NY NOAA` data

```{r message = FALSE}
library(p8105.datasets)
library(tidyverse)

data("ny_noaa")

noaa_obs = nrow(ny_noaa)

summary(ny_noaa)
```

Short description: This dataset contains `noaa_obs` weather observations, each identified by a unique station ID and date ranging from 1981 to 2010. Key variables include precipitation (`prcp`), snowfall (`snow`), and snow depth (`snwd`), along with maximum (`tmax`) and minimum (`tmin`) temperatures, though the temperature variables are recorded as characters. The dataset shows considerable missing data, with 145,838 missing values for precipitation, 381,221 for snowfall, and 591,786 for snow depth, indicating the need for careful handling of missing data in any analysis.

## Convert date to Date type and create separate variables for year, month, and day

```{r}
ny_noaa <- ny_noaa |>
  mutate(
    date = as.Date(date, format = "%Y-%m-%d"),
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )|>
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin)
  )
```

# Examine the most commonly observed values for snowfall

```{r}
snowfall_counts <- ny_noaa |>
  filter(!is.na(snow)) |>
  count(snow) |>
  arrange(desc(n))

print(head(snowfall_counts, 1))
```

The most commonly observed values for snowfall are 0 mm, indicating that the majority of days in the data set did not have any snowfall. This is expected, as snowfall is generally a rare event, especially outside of winter months. 

## Plot average max temperature in January and July across years for each station

```{r average max temperature plot by station}
avg_max_temp_plot <- ny_noaa |>
  filter(month %in% c(1, 7)) |>
  group_by(year, month, id) |>
  summarize(avg_tmax = mean(tmax, na.rm = TRUE)) |>
  mutate(month = ifelse(month == 1, "January", "July")) |>
  ggplot(aes(x = year, y = avg_tmax, color = id)) +
  geom_line() +
  facet_grid(.~ month) +
  labs(
    title = "Average Maximum Temperature in January and July Across Years by Station",
    x = "Year",
    y = "Average Maximum Temperature (°C)",
    color = "Station ID"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Display the plot
print(avg_max_temp_plot)
```

The mean temperature in January is much lower than the mean temperature in July for all stations and across all years. We see one extremely cold station in July of 1980s, as well as a few other less extreme outliers.

## Create a two-panel plot

## (i) tmax vs tmin for the full dataset using a hexbin plot
```{r}
hexbin_plot <- ny_noaa |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(
    title = "Hexbin Plot of Tmax vs Tmin",
    x = "Minimum Temperature (°C)",
    y = "Maximum Temperature (°C)"
  ) +
  theme_minimal()

```

## (ii) Distribution of snowfall values greater than 0 and less than 100 by year

```{r}
library(ggridges)

snowfall_distribution_plot <- ny_noaa |>
  filter(snow > 0 & snow < 100) |>
  ggplot(aes(x = snow, y = factor(year))) +
  geom_density_ridges() +
  labs(
    title = "Distribution of Snowfall Values (> 0 and < 100 mm) by Year",
    x = "Year",
    y = "Snowfall (mm)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r two-panel plot}
library(patchwork)

combined_plot <- hexbin_plot + snowfall_distribution_plot

print(combined_plot)
```

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution.

From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and  35 mm of snow in a year. Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm.

# Problem 2

## Imort data sets

```{r message = FALSE}
library(tidyverse)

nhanes_accel <- read_csv("hw3data/nhanes_accel.csv") |>
  janitor::clean_names() |>
  pivot_longer(cols = min1:min1440, names_to = "minute", values_to = "mims") |>
  mutate(minute = as.numeric(gsub("min", "", minute)))
  
nhanes_covar <- read_csv("hw3data/nhanes_covar.csv", skip = 4)
colnames(nhanes_covar) <- c("seqn", "sex", "age", "bmi", "education")
```

## Encode data with reasonable variable classes; Filter out participants less than 21 years of age and those with missing demographic data in `nhanes_covar`data set

Here we assume for `sex`, "1" is Male and "2" is female;
And we assume for `education`, "1" is Less than high school, "2" is High school equivalent and "3" is More than high school

```{r}
nhanes_covar <- nhanes_covar |>
  mutate(
    seqn = as.integer(seqn),
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    age = as.numeric(age),
    bmi = as.numeric(bmi),
    education = factor(education, levels = c(1, 2, 3), 
                       labels = c("Less than high school", 
                                  "High school equivalent", 
                                  "More than high school"))
  ) |>
  filter(age >= 21) |>
  drop_na()
```

## Merge the two data sets

```{r}
merged_data <- nhanes_accel |>
  inner_join(nhanes_covar, by = "seqn")
```

## Produce a reader-friendly table for the number of men and women in each education category

```{r}
nhanes_covar |>
  count(sex, education) |>
  pivot_wider(names_from = sex, values_from = n) |>
  knitr::kable()
```

## Create a visualization of the age distributions for men and women in each education category

```{r age distributions}
age_dist_plot <- nhanes_covar |>
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.3) +
  facet_grid(.~ education) +
  labs(
    title = "Age Distributions for Men and Women by Education Category",
    x = "Age",
    y = "Density",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(age_dist_plot)
```

Some comments:

* For the reader_friendly table:

   * There is a roughly equal distribution between men and women for the "Less than high school" group.

   * For "High school equivalent", there are more men than women.

   * In the "More than high school" group, the numbers of men and women are almost the same, with a slight excess of females.

* For the plot:

  * Less than high school: Both men and women have a median age around 60, with a roughly equal distribution.

  * High school equivalent: Men tend to have a lower median age of around 53, while the median age for women is over 60.

  * More than high school: The median age for men is around 42, while women are slightly younger (around 41). For 20-40 age group, women have a significantly greater density than men.

## Plot total activities (y-axis) against age (x-axis), comparing men to women, with separate panels for each education level

```{r total activities against age}
act_age_plot <- merged_data |>
  group_by(seqn, age, education, sex) |>
  summarize(total_activity = sum(mims)) |>
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  facet_grid(.~ education) +
  labs(
    title = "Total Activity vs Age by Education Level and Sex",
    x = "Age",
    y = "Total Activity",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(act_age_plot)
```

Some comments:

* In the "Less than high school" group, physical activity tends to decrease with age for both men and women, with some fluctuations in later adulthood (50-60). Men and women exhibit similar trends, but there are noticeable individual variations.

* In the "High school equivalent" group, women generally show higher total activity compared to men, especially in middle age (30-50). Activity levels for both sexes tend to decrease after 40 years old but increase before 40 years old.

* In the "More than high school" group, the trends are relatively stable across age, with men and women showing similar activity patterns, though women tend to have higher activity level. Activity levels for both sexes tend to decrease after 60 years old.

## Plot 24-hour activity time courses for each education level, using color to indicate sex

```{r 24-hour activity time courses}
day_course_plot <- merged_data |>
  ggplot(aes(x = minute, y = mims, group = seqn, color = sex)) +
  geom_line(alpha = 0.5) +
  geom_smooth(aes(group = sex), se = FALSE) +
  facet_grid(.~ education) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level and Sex",
    x = "Minute of the Day",
    y = "Mean Activity",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(day_course_plot)

```

Some comments:

* Less than high school: Both men and women show distinct troughs in activity in the early morning (250 mins~4 a.m.) and peaks in the afternoon (750 mins~12:30 p.m.). After 12:30 p.m., both men and women keep decreasing until night.

* High school equivalent: Both men and women show distinct troughs in activity in the early morning (250 mins~4 a.m.) and peaks in the afternoon (750 mins~12:30 p.m.). The peak will last until about (1125 mins~6:45 p.m.). After 6:45 p.m., both men and women keep decreasing until night.

* More than high school: The high education has the similar trend as the median education. However, women show a higher activity level than men during the peak time period. Both male and female show some significant outliers.

# Problem 3

## Import, clean, and tidy these data, and describe the resulting dataset.

```{r message = FALSE}
library(tidyverse)

jan_2020 <- read_csv("hw3data/Jan 2020 Citi.csv") |>
  mutate(month = "jan", year = 2020) |>
  janitor::clean_names()
jan_2024 <- read_csv("hw3data/Jan 2024 Citi.csv") |>
  mutate(month = "jan", year = 2024) |>
  janitor::clean_names()
jul_2020 <- read_csv("hw3data/July 2020 Citi.csv") |>
  mutate(month = "jul", year = 2020) |>
  janitor::clean_names()
jul_2024 <- read_csv("hw3data/July 2024 Citi.csv") |>
  mutate(month = "jul", year = 2024) |>
  janitor::clean_names()

combined_data <- bind_rows(jan_2020, jul_2020, jan_2024, jul_2024) |>
  distinct()

obs = nrow(combined_data)

summary(combined_data)
```

Description: The cleaned Citi Bike data set contains `r obs` observations as well as information on rides taken during January and July of both 2020 and 2024. It includes data such as the ride ID, bike type, day of the week, ride duration, start and end station names, as well as whether the rider is a member or casual user. The data set has been cleaned by combining all four months, removing duplicates, and standardizing column names.

## Produce a reader-friendly table

```{r}
combined_data |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n()) |>
  pivot_wider(names_from = member_casual, values_from = total_rides) |> 
  knitr::kable()
```

Some comments: The summary table shows the total number of rides taken by casual riders and Citi Bike members across different years and months. From the table, it is evident that there is a general trend of higher ridership in July compared to January, which can be attributed to better weather conditions during summer months. Additionally, Citi Bike members consistently take more rides than casual riders across all time periods, highlighting the benefit of membership for frequent users. Notably, the number of casual riders has increased from 2020 to 2024.

## Create a table showing the 5 most popular starting stations for July 2024

```{r}
combined_data |>
  filter(year == 2024, month == "jul") |>
  group_by(start_station_name) |>
  summarize(total_rides = n()) |>
  arrange(desc(total_rides)) |>
  slice_head(n = 5) |>
  knitr::kable()
```

From the summary table, Pier 61 at Chelsea Piers, University Pl & E 14 St, W 21 St & 6 Ave, West St & Chambers St, W 31 St & 7 Ave are the 5 most popular starting stations for July 2024.

## The effects of day of the week, month, and year on median ride duration

```{r median ride duration}
median_ride_plot <- combined_data |>
  mutate(weekdays = factor(weekdays, 
                           levels = c("Monday", "Tuesday", "Wednesday", "Thursday",                                              "Friday", "Saturday", "Sunday"))) |>
  group_by(year, month, weekdays) |>
  summarize(median_ride = median(duration, na.rm = TRUE)) |>
  ggplot(aes(x = weekdays, y = median_ride, color = interaction(year, month))) +
  geom_line(aes(group = interaction(year, month))) +
  facet_grid(year ~ month) +
  labs(
    title = "Median Ride Duration by Day of the Week, Month, and Year",
    x = "Day of the Week",
    y = "Median Ride Duration (minutes)",
    color = "Year.Month"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom")

print(median_ride_plot)
```

Some comments:

* In July (summer), median ride duration is generally longer compared to January (winter).

* Comparing 2024, 2020 has the noticeable higher median ride duration.

* Ride duration tends to be longer on weekends.

## Impact of month, membership status, and bike type on the distribution of ride duration

```{r distribution of ride duration}
ride_distribution_plot <- combined_data |>
  filter(year == 2024) |>
  ggplot(aes(x = rideable_type, y = duration)) +
  geom_violin() +
  facet_grid(member_casual ~ month) +
  labs(
    title = "Ride Duration Distribution by Month, Membership Status, and Bike Type (2024)",
    x = "Ride Duration (minutes)",
    y = "Count",
    color = "Member/Casual"
  ) +
  theme_minimal()  +
  theme(legend.position = "bottom") 

print(ride_distribution_plot)
```

Some comments:

* Casual riders appear to have a wider spread in ride duration compared to members, indicating more variability in their ride habits.

* July shows significantly more rides compared to January, which may indicate the influence of warmer weather encouraging more riding activity. 

* Electric bikes generally show a slightly higher concentration of shorter rides compared to classic bikes.
